Combining neural networks and geostatistics for landslide hazard assessment of San Salvador Metropolitan Area, El Salvador \newline \newline Combinando redes neuronales y geoestad\'{i}stica para evaluaci\'{o}n de deslizamientos de tierra del \'{a}rea Metropolitana de San Salvador, El Salvador

### 2015-09-28 paper.tex

\section{Introduction}
\label{sec:intr}
El Salvador, one of the smallest and most crowded nations in Central America, extends in Pacific coast about 240 km westward from the Gulf of Fonseca to the border with Guatemala (see Fig. \ref{fig:mass01}). El Salvador borders an active subduction boundary between Cocos and Caribbean plates located 30 km offshore. Therefore El Salvador is affected with high seismicity and volcanic activity related to this active boundary. There are two main sources of seismic activity: the upper interface thrust that coincides with the location of the recent volcanoes and the Benioff-Wadaty zones of subducted Cocos plate where deeper intraplate earthquakes occurs \cite{dewey}. Due this tectonic and volcanic activity El Salvador has rugged relief. The main ranges cross the country with a rough west-east trend, parallel to the coastline. These are separated from each other by faults and grabens. These ranges present several highly active volcanoes. The surface geology in El Salvador is almost entirely volcanic, dominated by upper Tertiary to Holocene volcanic rocks. Only sparse outcrops of sedimentary and plutonic rocks are located in the northern ranges, in the border with Honduras \cite{weber}. Some of the most recent volcanic layers are formed by poorly consolidated ashes and highly erodible tuffs \cite{bommer}.

@@

Popular classification models have been used such as logistic regression (\cite{akgun2012}, \cite{gaskill}, and \cite{garcia2008}), neural networks (\cite{garcia2010}, \cite{Melchiorre2011410},\cite{Zeng2001374}, \cite{Ermini2005327}, and \cite{Yesilnacar2005251}), and support vector machines (\cite{ballabio2012support} and \cite{tien2012landslide}).

@@

\section{Study area}
\label{sec:studyarea}
MASS is formed by 14 municipalities among them is located San Salvador, the capital of the country. It has 587 $km^{2}$ and according to the most recent census \cite{minecon} it has 1,566,629 inhabitans with a density of population of 2669 hab/km2. Geographically, MASS is extended over a flat erosional surface (650-760 m.a.s.l.), gently sloping to the east towards Ilopango Caldera also known as Ilopango Lake. MASS is bordered to the south by the Balsamo Range and San Jacinto Hill, to the west by San Salvador volcano and Mariona Hills (refer to Fig. \ref{fig:mass01}). Human settlement is not restricted to the plain and it spreads up to surrounding heights and volcano flanks.

All rocks exposed in MASS have volcanic origin and consists of intercaled primary and reworked deposits from Late Tertiary to Holocene age \cite{schmidt1975}. The younger ones are known as Tierra Blanca pyroclastic ash deposits and forms a relative continuous layer in the majority of MASS with an average thickness of 4 m \cite{schmidt1975}. These are deposits derived mainly from Ilopango caldera eruptions where they are more than 50 m of thick \cite{schmidt1975}. Tierra Blanca soils is highly susceptible to erosion, one of the most common dangerous effects of the erosion in this area are landslides during heavy rainfall and earthquake ground shaking \cite{schmidt1975}. Ch\'{a}vez et al. \cite{chavez2014a} identified high and moderate erosion hazard in the majority of MASS area. The more recent layer Tierra Blanca, known as Tierra Blanca Joven (TBJ) \cite{hernan2004} presents very poor geotechnical properties, especially when there is an increment of soil moisture (i.e. during rainy seasons). This results in high instability susceptibility (\cite{chavez2014b} and \cite{rolo2004}). Several incised rivers and ravines cross MASS area. Due to the large erosion rates of the area, natural slopes of these watercourses are often close to vertical and can reach highs of over 10 m. Taking into account this property of volcanic soils, usually vertical slopes are cut to urban and road construction. As Bommer \& Rodriguez \cite{bommer} identified in Central America, although such slopes may remain stable for years, they may become unstable, abruptly and totally under the action of heavy rainfall or seismic shaking. Hern\'{a}ndez \cite{hernan2004} carried out a detailed description of this process in TBJ, where these instabilities are very common. 
In the study area fault trends are characterized by steep dip angles (65\textdegree - 90\textdegree). The main fault system is characterized by an east-west trend; this trend is responsible for the steep northern slope of the El Balsamo Range.  The secondary systems are with north, northwest and northeast trends offset the main fault system and are more active in the present tectonic setting \cite{schmidt1975}. However, all the mentioned fault families appear active. Thus the several fault systems have developed at different times, but been repeatedly activated \cite{rymer1987}. Finally there are also several ringlike faults that can be related to subsurface collapse due to volcanic tectonic subsidence \cite{schmidt1975}. The Last eruption of San Salvador volcano was in 1917.  The majority of the magma's total volume (97\%) extruded from several vents in the Northwest flank of the volcano, outside of MASS. However a fraction of MASS is located under the proximal volcanic hazard zone according to Sofield \cite{sofield2004}. In the steep slopes  San Salvador volcano, especially of Picacho, the highest peak of the volcano, located in the eastern side presents a high lahar hazard that threats MASS \cite{major2004}. On the other hand one of the most destructive eruptions in the world history was 430 A.D. TBJ eruption produced by the Ilopango Caldera. According to Dull \cite{dull2004}, that eruption produced total devastation in whole MASS area.

@@

In the equation (\ref{eq:ann3}) $p_{i}$ is the the predicting probability for a network with $ k^{*} $ input characteristics and $ j^{*} $ neurons. In the context of the present paper, $p_{i} $ represents the probability of landslide ocurrence.

@@

The method used for estimating the parameters of the model was an Hybrid Method \cite{McNelis2005}. Firstly, a heuristic genetic algorithm, using a package developed in the R statistical software \cite{rproject} specifically developed for this purpose, was used to obtain a good estimation of the parameters of the model. The R package can be accessed from the following Web address:

@@

Once a good estimate was obtained, this was used as the initial guess value for the conjugate gradient method implemented in the R statistical software (optim function), to obtain a better estimation of the parameters of the model.

@@

In standard statistical problems, correlation can be estimated from a scatterplot, when several data pairs ${x, y}$ are available. The spatial correlation between two observations of a variable $z(s)$ at locations $s_{1}$ and $s_{2}$ cannot be estimated because only a single pair is available. To estimate spatial correlation
from observational data, it is necessary to make stationarity assumptions
before we can make any progress. One commonly used form of stationarity
is intrinsic stationarity, which assumes that the process that generated the
samples is a random function $Z(s)$ composed of a mean and residual \cite{bivand2008applied}:

@@

Regarding variogram estimation, the spherical, gaussian and exponential models were proposed. Using cross-validation methodology, the best model was the exponential, since it had an $ R^{2} = 0.42 $ against $ R^{2} = 0.40 $ and $ R^{2} = 0.39 $ of the models spherical and gaussian, respectively. Using exponential model, spatial predictions were calculated on a grid of 1500 $\times$ 1500 and exported to Geotiff raster format. Finally, Geographic Information System (ArcGIS\textregistered \  of ESRI) was used to design the final map (Figure \ref{fig:mass02}). This map (Fig. \ref{fig:mass02}) represents the landslide probability of occurrences in MASS. According to Fig. \ref{fig:mass02} higher probability of landslide occurrence are around 0.8. These are located in the northern slopes of Balsamo Range (western side of MASS), in the eastern slope of San Salvador volcano and Picacho,  along the extension of Mariona Hills  and in the southeastern side, between Ilopango lake and San Jacinto Hill. Areas with lower probability, but still important, are located in southern and western side of Nejapa Hill, in north-east of the studied area, in northern banks of Ilopango lake, in northeast slope of San Salvador Volcano, in the surrounding areas of San Jacinto Hill, and in general in southern side of MASS.

@@

The results obtained through geostatistical methodology met the assumptions for kriging method application. Final map of landslide probability of occurrence (see Fig. \ref{fig:mass02} ) generated by kriging method was the result of multidisciplinary work which implied an intense geological analysis (i.e. MARN inventory of 4792 landslides), together with the use of sophisticated statistical techniques. This landslide hazard map is a user-friendly tool to evaluate landslide risk in MASS and can be useful for many purposes such as territorial planning, prevention and landslide mitigation and so on. 
\begin{center}
  \begin{figure}
   \centering
   \includegraphics[scale=0.73]{MASS_map_2.eps}
   \caption{\small{Landslide hazard map MASS where the landslide probability of occurrence is shown.}}
   \label{fig:mass02}
  \end{figure}
 \end{center}
High probability of landslide occurrence in Picacho slope implies very high risk due to the proximity of urban areas and due to the large volume of mobilized materials. Major et al.\cite{major2004} identified in San Salvador volcano lahars as voluminous as 2 million of $m^{3}$ that can affect large zones of urban areas. Southern areas of MASS also have a high risk because landslides, similar to Las Colinas landslide which involved a total volume of around 200,000 $m^{3}$ \cite{evans}, can occur in northern slope of Balsamo range and San Jacinto Hill, affecting very crowded urban areas.

The present hazard landslide map is consistent with former landslide hazard maps carried out in the region using approaches conceptually different. Fern\'{a}ndez-Lavado et al.\cite{fernan2008} presented a landslide hazard GIS map of MASS generated through bivariate statistical method considering several conditioning factors showing similar hazard areas. The only differences are located mainly in the Nejapa Hill and San Jacinto Hill where Fern\'{a}ndez-Lavado et al. \cite{fernan2008} identified a slightly higher hazard and the area between San Jacinto Hill and Ilopango Lake where the former study presented a hazard slightly lower. Since landslides are an erosional phenomenon, the highest probability areas of landslide occurrence coincide with the areas with highest hazard of erosion in the erosion hazard map of MASS proposed by Ch\'{a}vez et al. \cite{chavez2014a}. There are similarities with susceptibility hazard maps at national scale generated through logistic regression method \cite{garcia2008}, through artificial neural network method \cite{garcia2010}, and through Mora-Vahrson method \cite{snet2004}. However the high landslide areas are more extended in these national maps than the present study. These differences could be explained by the working scale. Our map is more detailed because is at regional scale. Other possible explanation can be related to the data. Present work was built using a more extended landslide catalog and more up-to-date data (e.g. seismic hazard evaluation from Benito et al. \cite{beni2012}) than former works.

Rock falls were not included in the catalog; these were very common during 2001 earthquakes in El Balsamo Range \cite{jibson}. On the other hand, the MASS seismic microzonation of seismic site effects will improve the estimation of peak ground acceleration. The local geologic conditions are responsible for the modifications (amplification, frequency content and duration) experimented by seismic motions just before reaching the ground surface. Additionally, the geometry of the ground surface also may produce amplification (topographic effects) \cite{aki1993}. These effects can be important in soft soils close to incised rivers and also in San Salvador volcano, Balsamo ranges and hills. Crosta et al. \cite{crosta2005} identified an amplification factor of 1.3-1.4 in the Las Colinas failure area (Balsamo Ranges). All these variables can improve the predictive ability of the model and the likelihood of landslides.

>>>>

\section{Introduction}
\label{sec:intr}
El Salvador, one of the smallest and most crowded nations in Central America, extends in Pacific coast about 240 km westward from the Gulf of Fonseca to the border with Guatemala (see Fig. \ref{fig:mass01}). El Salvador has an active subduction boundary between Cocos and Caribbean plates located 30 km offshore. Therefore El Salvador is affected with high seismicity and volcanic activity related to this active boundary. There are two main sources of seismic activity: the upper interface thrust that coincides with the location of the recent volcanoes and the Benioff-Wadaty zones of subducted Cocos plate where deeper intraplate earthquakes occurs \cite{dewey}. Due this tectonic and volcanic activity El Salvador has rugged relief. The main ranges cross the country with a rough west-east trend, parallel to the coastline. These are separated from each other by faults and grabens. These ranges present several highly active volcanoes. The surface geology in El Salvador is almost entirely volcanic, dominated by upper Tertiary to Holocene volcanic rocks. Only sparse outcrops of sedimentary and plutonic rocks are located in the northern ranges, in the border with Honduras \cite{weber}. Some of the most recent volcanic layers are formed by poorly consolidated ashes and highly erodible tuffs \cite{bommer}.

@@

Popular classification models have been used, such as logistic regression (\cite{akgun2012}, \cite{gaskill}, and \cite{garcia2008}), neural networks (\cite{garcia2010}, \cite{Melchiorre2011410},\cite{Zeng2001374}, \cite{Ermini2005327}, and \cite{Yesilnacar2005251}), and support vector machines (\cite{ballabio2012support} and \cite{tien2012landslide}).

@@

\section{Study area}
\label{sec:studyarea}
MASS is formed by 14 municipalities among them is San Salvador, the capital of the country. It has 587 $km^{2}$ and according to the most recent census \cite{minecon} it has 1,566,629 inhabitans with a density of population of 2669 hab/km2. Geographically, MASS is extended over a flat erosional surface (650-760 m.a.s.l.), gently sloping to the east towards Ilopango Caldera also known as Ilopango Lake. MASS is bordered to the south by the Balsamo Range and San Jacinto Hill, to the west by San Salvador volcano and Mariona Hills (refer to Fig. \ref{fig:mass01}). Human settlement is not restricted to the plain and it spreads up to surrounding heights and volcano flanks.

All rocks exposed in MASS have volcanic origin and consist of intercaled primary and reworked deposits from Late Tertiary to Holocene age \cite{schmidt1975}. The younger ones are known as Tierra Blanca pyroclastic ash deposits and forms a relative continuous layer in the majority of MASS with an average thickness of 4 m \cite{schmidt1975}. These are deposits derived mainly from Ilopango caldera eruptions with more than 50 m of thick \cite{schmidt1975}. Tierra Blanca soils is highly susceptible to erosion, one of the most common dangerous effects of the erosion in this area are landslides during heavy rainfall and earthquake ground shaking \cite{schmidt1975}. Ch\'{a}vez et al. \cite{chavez2014a} identified high and moderate erosion hazard in the majority of MASS area. The more recent layer Tierra Blanca, known as Tierra Blanca Joven (TBJ) \cite{hernan2004} presents very poor geotechnical properties, especially when there is an increment of soil moisture (i.e. during rainy seasons). This results in high instability susceptibility (\cite{chavez2014b} and \cite{rolo2004}). Several incised rivers and ravines cross MASS area. Due to the large erosion rates of the area, natural slopes of these watercourses are often close to vertical and can reach highs of over 10 m. Taking into account this property of volcanic soils, usually vertical slopes are cut to urban and road construction. As Bommer \& Rodriguez \cite{bommer} identified in Central America, although such slopes may remain stable for years, they may become unstable, abruptly and totally under the action of heavy rainfall or seismic shaking. Hern\'{a}ndez \cite{hernan2004} carried out a detailed description of this process in TBJ, where these instabilities are very common. 
In the study area fault trends are characterized by steep dip angles (65\textdegree - 90\textdegree). The main fault system is characterized by an east-west trend; this trend is responsible for the steep northern slope of the El Balsamo Range.  The secondary systems are with north, northwest and northeast trends offset the main fault system and are more active in the present tectonic setting \cite{schmidt1975}. However, all the mentioned fault families appear active. Thus the several fault systems were formed at different times, but were repeatedly reactivated \cite{rymer1987}. Finally there are also several ringlike faults that can be related to subsurface collapse due to volcanic tectonic subsidence \cite{schmidt1975}. The last eruption of San Salvador volcano was in 1917.  The majority of the magma's total volume (97\%) extruded from several vents in the Northwest flank of the volcano, outside of MASS. However a fraction of MASS is located under the proximal volcanic hazard zone according to Sofield \cite{sofield2004}. In the steep slopes of San Salvador volcano, especially at Picacho, the highest peak which is located in the eastern side, there is a high lahar hazard that threatens MASS \cite{major2004}. On the other hand one of the most destructive eruptions in the world history was 430 A.D. The TBJ eruption was produced by the Ilopango Caldera and  according to Dull \cite{dull2004}, yielded total devastation in whole MASS area.

@@

In the equation (\ref{eq:ann3}) $p_{i}$ is the output of a neural network with $ k^{*} $ input characteristics and $ j^{*} $ neurons. In the context of the present study, $p_{i} $ represents the probability of landslide ocurrence.

@@

The method used for estimating the parameters of the model was an Hybrid Method \cite{McNelis2005}. Firstly, a genetic algorithm was implemented in a package developed in the R statistical software \cite{rproject} to obtain a good estimation of the parameters of the model. The R package can be accessed from the following Web address:

@@

After obtaining a good estimate by genetic algorithms, this was used as initial guess value for the conjugate gradient method implemented in the R statistical software (optim function), to obtain a better estimation of the parameters of the model.

@@

In standard statistical problems, correlation can be estimated from a scatterplot, when several data pairs ${x, y}$ are available. The spatial correlation between two observations of a variable $z(s)$ at locations $s_{1}$ and $s_{2}$ cannot be estimated because only a single pair is available. To estimate spatial correlation
from observational data, it is necessary to make stationarity assumptions
before we can make any progress. One commonly used form of stationarity
is intrinsic stationarity, which assumes that the process that generated the
samples is a random function $Z(s)$ composed of a mean and residual \cite{bivand2008applied} as follows:

@@

Regarding variogram estimation, the spherical, gaussian and exponential models were proposed. Using cross-validation methodology, the best model was the exponential, since it had an $ R^{2} = 0.42 $ against $ R^{2} = 0.40 $ and $ R^{2} = 0.39 $ of the models spherical and gaussian, respectively. Using exponential model, spatial predictions were calculated on a grid of 1500 $\times$ 1500 and exported to Geotiff raster format. Finally, Geographic Information System (ArcGIS\textregistered \  of ESRI) was used to design the final map (Figure \ref{fig:mass02}). This map (Fig. \ref{fig:mass02}) represents the landslide probability of occurrences in MASS. According to Fig. \ref{fig:mass02} higher probability of landslide occurrence are around 0.8. These are located in the northern slopes of Balsamo Range (western side of MASS), in the eastern slope of San Salvador volcano and Picacho,  along the extension of Mariona Hills  and in the southeastern side, between Ilopango lake and San Jacinto Hill. Areas with lower probability are located in southern and western side of Nejapa Hill, in north-east of the studied area, in northern banks of Ilopango lake, in northeast slope of San Salvador Volcano, in the surrounding areas of San Jacinto Hill, and in general in southern side of MASS.

@@

The results obtained through geostatistical methodology met the assumptions for kriging method application. Final map of landslide probability of occurrence (see Fig. \ref{fig:mass02} ) generated by kriging method was the result of multidisciplinary work which implied an intense geological analysis (i.e. MARN inventory of 4792 landslides), together with the use of sophisticated statistical techniques. This landslide hazard map is a user-friendly tool to evaluate landslide risk in MASS and can be useful for many purposes such as territorial planning, prevention and landslide mitigation and so on. 
\begin{center}
  \begin{figure}
   \centering
   \includegraphics[scale=0.73]{MASS_map_2.eps}
   \caption{\small{Landslide hazard map MASS where the landslide probability of occurrence is shown.}}
   \label{fig:mass02}
  \end{figure}
 \end{center}
High probability of landslide occurrence in Picacho slope implies very high risk due to the proximity of urban areas and the large volume of mobilized materials. Major et al.\cite{major2004} identified in San Salvador volcano lahars as voluminous as 2 million of $m^{3}$ that can affect large zones of urban areas. Southern areas of MASS also have a high risk of landslides that is similar to Las Colinas landslide which involved a total volume of around 200,000 $m^{3}$ \cite{evans}. A similar risk exists in northern slope of Balsamo range and San Jacinto Hill, affecting very crowded urban areas.

The present hazard landslide map is consistent with former landslide hazard maps carried out in the region using approaches conceptually different. Fern\'{a}ndez-Lavado et al.\cite{fernan2008} presented a landslide hazard GIS map of MASS generated through bivariate statistical method considering several conditioning factors showing similar hazard areas. The only differences are located mainly in the Nejapa Hill and San Jacinto Hill where Fern\'{a}ndez-Lavado et al. \cite{fernan2008} identified a slightly higher hazard and the area between San Jacinto Hill and Ilopango Lake where the former study presented a hazard slightly lower. Since landslides are an erosional phenomenon, the highest probability areas of landslide occurrence coincide with the areas with highest hazard of erosion in the erosion hazard map of MASS proposed by Ch\'{a}vez et al. \cite{chavez2014a}. There are similarities with susceptibility hazard maps at national scale generated through logistic regression \cite{garcia2008}, artificial neural network \cite{garcia2010}, and Mora-Vahrson \cite{snet2004}. However the high landslide areas are more extended in these national maps than the present study. These differences could be explained by the working scale. Our map is more detailed because is at regional scale. Other possible explanation can be related to the data. Present work was built using a more extended landslide catalog and more up-to-date data (e.g. seismic hazard evaluation from Benito et al. \cite{beni2012}) than former works.

Rock falls were not included in the catalog; these were very common during 2001 earthquakes in El Balsamo Range \cite{jibson}. On the other hand, the seismic microzonation of MASS will improve the estimation of peak ground acceleration. The local geologic conditions are responsible for the modifications (amplification, frequency content and duration) experimented by seismic motions just before reaching the ground surface. Additionally, the geometry of the ground surface also may produce amplification (topographic effects) \cite{aki1993}. These effects can be important in soft soils close to incised rivers and also in San Salvador volcano, Balsamo ranges and hills. Crosta et al. \cite{crosta2005} identified an amplification factor of 1.3-1.4 in the Las Colinas failure area (Balsamo Ranges). All these variables can improve the predictive ability of the model and the likelihood of landslides.
